---
title: "An Analysis of Red Wine"
author: "Guy Bourne"
date: "28 November 2017"
output: html_document
---
*Built with R `r getRversion()`*

## Introduction

Let me start with a disclaimer, I drink more wine than I should and less than I
would like. Since moving to Australia, I have been spoilt for choice with a 
range of fantastic Aussie wines and when I realised that I had the opportunity 
to understand a little more about them whilst learning some new skills, well 
who could resist? 

> *"After the first glass, you see things as you wish they were. After the*
*second, you see things as they are not. Finally, you see things as they really*
*are, and that is the most horrible thing in the world."* 
                                                         **Oscar Wilde**
                                                         
Given Oscar Wilde's warning, this analysis will be conducted sober, but the hope 
is that we will see the world of Red Wine as it really is, and that it won't be 
all that bad, plus we'll have some fun getting there.

Note that throughout this document, I'll be experimenting with different 
aesthetics, so I am not aiming for consistency across plots; this would drive
me crazy as a reviewer so apologies upfront!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

## Install these if necessary

# options(repos=c(CRAN='cran.rstudio.com'))

# install.packages("factoextra")
# install.packages("FactoMineR")
# install.packages("dplyr")
# install.packages("ggthemes")
# install.packages('reshape2')
# install.packages('GGally')
# install.packages('scales')
# install.packages('memisc')
# install.packages('lattice')
# install.packages('MASS')
# install.packages('car')
# install.packages('reshape')
# install.packages('plyr')
# install.packages("ggplot2")
# install.packages('kableExtra')
# install.packages('gridExtra')

# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(knitr)
library(kableExtra)
library(reshape2)
library(gridExtra)
library(ggthemes)
library(dplyr)
library(FactoMineR)
library(factoextra)

## Set some global options
options(knitr.table.format = "html") 
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('C:\\Users\\geeky\\Dropbox\\Learning\\MOOCs\\Udacity\\Data Analysis NanoDegree\\EDA\\Project\\Udacity-DataAnalysis-EDA')

redWine <- read.csv(file = 'wineQualityReds.csv')
redWine$X <- factor(redWine$X)

```

We'll be exploring the measurable aspects of red wine and seeing how they 
interact, looking for relationships, whilst also seeing how they relate to 
critics score of 1 - 10. Note that the id variable X has been converted into
a factor prior to analysis.

## Initial Univariate analysis

We'll start by summarising the dataset and producing a broad spectrum of plots
to introduce the data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#High level summaries of the data
str(redWine)
```

Firstly, we note that there are no categorical variables in this dataset. 
This may limit my ability to learn more about R plotting, however I think that 
we'll be able to create some categorical buckets later to make sure that we can
add some extra dimensions to plots.

```{r echo = FALSE, message=FALSE, warning=FALSE}

## Create basic univariate statistics for all vars 

for (i in  names(redWine)) {
  if (i != 'X') {
    assign(paste(i),summary(redWine[, i]))
  }
}

## Present in a table using kable to make it pretty

redWine.summary <- rbind(alcohol,
                         chlorides,
                         citric.acid,
                         density,
                         fixed.acidity,
                         free.sulfur.dioxide,
                         pH,
                         residual.sugar,
                         sulphates,
                         total.sulfur.dioxide,
                         volatile.acidity,
                         quality)

kable(redWine.summary) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_footnote(c(
    "All acids, sugars, sulphates and density are measured in g/dm^3",
    "Chlorides and sulphur dioxides are measured in mg/dm^3",
    "Alcohol is measured in % by volume",
    "Quality is a score between 1 and 10"), notation = "symbol")

```

I'm most interested in how much fun the judges had rating the wine, so 
I'd like to see the distribution of the strength of the wine (alcohol).

```{r echo=FALSE, fig.align = "center", message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol), data = redWine) +
  geom_histogram() + 
  theme_minimal()
```

What I find most interesting about this initial plot is that the distribution 
has a skew to the right. Living in Australia I'm used to quite bold red wines,
in particular Shiraz. In doing a little research I found a quote from James 
Halliday, a renowned wine expert.

> *"For the sake of simplicity, the alcohol of any red wine in excess of 14.5%
(alcohol by volume) can be said to be high. As recently as the 1980s, the 
benchmark was 13.5%. Ten years earlier, it might have been 12.5%."*
                                                         **James Halliday**

Note that we only have `r nrow(subset(redWine,redWine$alcohol >= 12.5))` wines
above the old standard, and just 
`r nrow(subset(redWine,redWine$alcohol >= 14.5))` 
above the more recent definition. This confirms that these are quite different 
from the new world wines that I tend to enjoy. It might be worth adding some 
indicators to split by these strength boundaries for later analysis.

Let's get on with it and create histograms for the other variables.

```{r echo=FALSE,fig.align = "center", message=FALSE, warning=FALSE}
## Create histograms for all variables (Excluding X)

ggplot(data = melt(redWine), mapping = aes(x = value)) + 
  geom_histogram(bins = 30) + facet_wrap(~variable, scales = 'free_x') + 
  theme_minimal()

## Add the strength boundaries for later
redWine$strength.category <- factor(
  ifelse(redWine$alcohol < 10 , 'Wk',
                                   ifelse(redWine$alcohol < 12.5, 'Mod',
                                          ifelse(redWine$alcohol < 14.5, 'Str'
                                                 , 'Aus')
                                   )
  ), levels = c('Wk','Mod','Str','Aus')
)

```

The sugar and sulphates have a left skew, which is to be expected given the bias
towards low alcohol wines in the data. I would have expected the acidity to have 
roughly the same shape, so that might be something worth considering as we 
start to look at bivariate relationships.

As expected the scale for quality follows a roughly normal distribution. The 
rough facet diagram is too granular, but we'll re-render it below.

```{r echo=FALSE,fig.align = "center", message=FALSE, warning=FALSE}

ggplot(aes(x=quality), data = redWine) +
  stat_count(geom = "bar") + 
  theme_minimal()

```

## Bivariate Plots Section

In this section I'm mostly interested in observing the relationship between 
quality. But before exploring that in detail, let's look at the relationships
between all the variables. Typically, I would start by looking at correlations 
and then expand this to scatter plots of all pairs of variables.


```{r echo=FALSE, Correlation_plot, fig.height = 12, fig.width = 12, fig.align = "center", message=FALSE, warning=FALSE}

ggcorr(redWine[, 2:13], geom = 'circle', min_size = 5, max_size = 15) + 
  theme_minimal()


```

Our strongest correlation with quality is the amount of alcohol, perhaps our 
judges were from the Barossa Valley! The level of citric acids and sulphates 
seems to have a moderate correlation, and we can see that having a large amount
of volatile acid can also annoy judges. This shouldn't come as a surprise, as 
this tends to be what can give wine a vinegar taste.

We'll plot all the scatters to see if anything jumps out, however given the 
number of variables it may be tricky to read the plots and we'll have to 
generate ones that look interesting and take a closer look later.

From this point on I'll be considering quality as a categorical variable, whilst 
it's been useful to calculate correlations; for multivariate plotting treating 
it as a factor should yield more insight.

```{r echo=FALSE, Bivariate_Plots, fig.height = 12, fig.width = 12,fig.align = "center", message=FALSE, warning=FALSE}

##Change quality to a factor

redWine$quality <- factor(redWine$quality)

ggpairs(redWine[, 2:13],mapping = aes(alpha = 0.05)) + 
  theme_minimal()


```

Let's explore some of the highly correlated variables in more detail. 

```{r echo = FALSE, violins1, fig.height = 12, fig.width = 15, fig.align = "center", message=FALSE, warning=FALSE}


ggplot(redWine, aes(quality,citric.acid)) + 
  geom_violin(trim = FALSE, fill='lightskyblue1', color="midnightblue") + 
  geom_boxplot(width = 0.1) + 
  theme_minimal()
```

This first violin plot reconfirms that better rated wines tend to have a higher 
concentration of citric acid. In the notes provided it states that this provides
a freshness and flavour to the wine. A little further research shows that citric
acid can be unstable and can ferment into acetic acid which is highly 
undesirable. Let's have a look at the relationship between citric acid and see
if the antimicrobial effect limits this risk. We had seen a positive correlation,
but let's explore further.

```{r echo = FALSE, scatter_citric, fig.align = "center", message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = sulphates), data = redWine) + 
  geom_point(aes(alpha = 0.2, color = quality)) +
  geom_smooth()

```

The default model applied here is a little misleading given what looks like an 
erroneous data point at (1,2). Let's try again, this time removing the outlier.
We'll also adjust the zoom a little to take a closer look at the trend without 
removing more of the data.

```{r echo = FALSE, scatter_citric2, fig.align = "center", message=FALSE, warning=FALSE}
ggplot(aes(x=citric.acid, y = sulphates), data = subset(redWine, 
                                                        redWine$citric.acid <1 )) + 
  geom_point(aes(color = quality)) +
  geom_smooth(color = 'black', fill = 'pink', method = lm)  +
  coord_cartesian(ylim = c(0.25, 1.25)) +
  scale_colour_grey() +
  theme_tufte()

```

This does still give us a positive trend, but it's not as significant as we
might have hoped. What we can see is that the quality increases with each 
variable. Let's look at the sulphates

```{r echo = FALSE, violins2, ,fig.height = 12, fig.width = 15, fig.align = "center", message=FALSE, warning=FALSE}
ggplot(redWine,aes(quality, sulphates)) + 
  geom_violin(trim = FALSE, fill = 'steelblue3', color = "midnightblue") + 
  geom_boxplot(width = 0.1) + 
  theme_minimal()
```

We see a similar trend again with a gradual increase in median values as 
we go up the quality grades. These plots show that the sulphates have less
volatility than the amount of citric acid in each category, and glancing back at
our first histogram plots we can see that it is a quite tight distribution with a 
standard deviation of `r sd(redWine$sulphates)` g/dm^3. Sulphates had quite a 
strong positive correlation with chlorides. Let's take a closer look.

```{r echo = FALSE, scatter_sulphates, fig.align = "center", message=FALSE, warning=FALSE}

ggplot(aes(x = sulphates, y = chlorides), data = redWine) + 
  geom_point(aes(color = quality)) +
  geom_smooth(color = 'black', fill = 'red', method = loess) +
  coord_cartesian(xlim = c(min(redWine$sulphates), 1.5)) +
  scale_colour_grey() +
  theme_economist()

```

Using the loess model to try and tease out a trend, we don't really find 
anything convincing and perhaps we should have stuck with a simple linear 
model. However, it doesn't tell us much more than the correlation matrix. 

```{r echo = FALSE, violins3, ,fig.height = 12, fig.width = 15, fig.align = "center", message=FALSE, warning=FALSE}
ggplot(redWine,aes(quality,alcohol)) + 
  geom_violin(trim = FALSE, fill = 'steelblue4', color = "midnightblue") + 
  geom_boxplot(width = 0.1) + 
  theme_minimal()
```

This plot is interesting, we had noted earlier that wines with higher alcohol 
content tended to get better scores, and whilst this is true, we can see that 
the standard deviation of alcohol content is different across the ratings.

```{r echo = FALSE, message=FALSE, warning=FALSE}

redWine.wine_by_quality <- redWine %>%
  group_by(quality) %>%
  summarise(al.stdDev = sd(alcohol),
            al.mean = mean(alcohol),
            al.median = median(alcohol),
            al.n = n()) %>%
  arrange(quality)

kable(redWine.wine_by_quality) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))


```

We can see that the standard deviation of the alcohol is increasing 
significantly as we go up the ratings. For our other two variables correlated
with quality, the distribution wasn't quite so wide.

## Multivariate Analysis

### Clusters

In this section we're going to cluster the data and observe the frequency
of ratings across the analytically derived clusters. Given R is new for me, I'm
going to stick to simple clustering with using k-means and I'll try a few values
of K.

```{r echo=FALSE, Multivariate_Plots, fig.align = "center", message=FALSE, warning=FALSE}

#Run the clustering algorithm

set.seed(1234)

km.res <- kmeans(scale(redWine[, 2:12]), 5, nstart = 25)

fviz_cluster(km.res, 
             data = redWine[, 2:12],
             geom = 'point',
             ggtheme = theme_minimal(),
             main = "kMeans Clustering Plot k=5"
)

```

This gives us a good starting point, we can see the 5 clusters across the two 
primary dimensions that have been created through principal components analysis
(PCA). Explaining PCA is beyond the scope of this exploration, however it is 
convenient for the presentation of multidimensional clusters. 

The whilst the groups do show separation, I think we could do better.

```{r echo=FALSE, Clustering2, fig.align = "center", message=FALSE, warning=FALSE}

#determine optimal cluster size

fviz_nbclust(scale(redWine[, 2:12]), kmeans, method = "gap_stat")

```

This tells us that to maximise the gap between our clusters we should 
produce a cluster diagram with 2 centroids. However, I think that having at least
3 would give us some more interesting talking points without sacrificing the
statistics. Let's look at the two cluster diagrams and make the call.

```{r echo=FALSE, Clustering3, fig.width = 15, fig.align = "center", message=FALSE, warning=FALSE}

#determine optimal cluster size


km.res.2 <- kmeans(scale(redWine[, 2:12]), 2, nstart = 25)
km.res.3 <- kmeans(scale(redWine[, 2:12]), 3, nstart = 25)

kmg.2 <- fviz_cluster(km.res.2, 
             data = redWine[, 2:12],
             geom = 'point',
             ggtheme = theme_minimal(),
             main = "kMeans Clustering Plot k=2"
)


kmg.3 <- fviz_cluster(km.res.3, 
             data = redWine[, 2:12],
             geom = 'point',
             ggtheme = theme_minimal(),
             main = "kMeans Clustering Plot k=3"
)

grid.arrange(kmg.2, kmg.3, ncol=2)

```

Whilst we have some overlap, I think we're ok to go with 3 clusters to made our
analysis a little more interesting. Let's add our clusters to the data and look 
at whether is related to our clusters.

```{r, echo=FALSE, fig.align = "center"}

## Bind the clusters to the original dataframe

redWine.clusters <- cbind(redWine, km.res.3[1])

## Convert clusters to text as we're treating as a category

redWine.clusters$cluster <- as.character(redWine.clusters$cluster)

## Plot it!

ggplot(aes(x=quality), data = redWine.clusters) +
  stat_count(geom = "bar", aes(fill = cluster)) + 
  theme_minimal()

## Tabulate it!


redWine.wine_by_cluster <- redWine.clusters %>%
  group_by(cluster) %>%
  summarise(qual.stdDev = sd(as.integer(quality)),
            qual.mean = mean(as.integer(quality)),
            qual.median = median(as.integer(quality)),
            qual.n = n()) %>%
  arrange(cluster)

kable(redWine.wine_by_cluster) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

It looks like we have a drift away from wines in cluster 2 and towards those in 
cluster 1. By changing the visualisation to represent proportions we can take
a closer look.

```{r, echo=FALSE, fig.align = "center"}

## Plot it!

ggplot(aes(x = quality), data = redWine.clusters) +
  stat_count(geom = "bar", aes(fill = cluster), position = 'fill') + 
  theme_minimal()

```

This has made the relationship a little clearer, however the question that 
springs to mind is, "ae the differences in proportion statistically 
significant?". Whilst we could model this with t-Tests or (even better) include
our clusters in a regression model to understand the importance, I fell that 
this is where our exploration will end for now. Time to wrap up!

------

## Final Plots and Summary

To summarise the findings, I'd like to showcase 3 plots. Firstly, 
the correlation matrix. I find this a hugely useful tool to declutter data,
particularly finding variables that may be equivalent or highly dependent. 
In this analysis it was primarily used to flag variables that seemed to linearly
influence the judges score.

### Plot One: Correlation Matrix
```{r echo=FALSE, Plot_One, fig.height = 12, fig.width = 12, fig.align = "center", message=FALSE, warning=FALSE}

# To recreate corr plot need quality as  numeric
redWine$quality <- as.numeric(redWine$quality)

ggcorr(redWine[, 2:13], geom = 'circle', min_size = 5, max_size = 15) + 
  labs(title = 'Red Wine Correlation Matrix', caption = 'Visualising linear 
       relationships between variables') + 
  theme_tufte()


```

### Description One
At a glance this plot tells us that alcohol, sulphates and citric acid are 
correlated with the judges' quality score. This makes sense. Our domain knowledge
and external research tells us that alcohol and citric acids can alter the 
bouquet and hence the taste. Sulphates act as a control preventing chemical 
reactions that can impair the wine.


### Plot Two: Violin plot of Alcohol distribution by quality
```{r echo=FALSE, Plot_Two, fig.height = 12, fig.width = 12, fig.align = "center", message=FALSE, warning=FALSE}

ggplot(redWine,aes(as.factor(quality), alcohol)) + 
  geom_violin(trim = FALSE, fill = 'steelblue4', color = "midnightblue") + 
  geom_boxplot(width = 0.1) + 
  labs(title = 'Violin plot of Alcohol Distribution by Quality',
       x = "quality",
       caption = '* Box plots represent median and IQR of distribution') +
  theme_tufte()


```

### Description Two
The main reason that I like this plot is that it can convey a lot of information
to an analyst quickly. Only by overlaying the distribution did it become 
apparent that the standard deviation of alcohol in highly rated wines was so 
broad, meaning that alcohol content alone cannot be considered to be a causal 
factor (which would not be the case if I were a judge, I'm sure.)  

### Plot Three: Clusters
```{r echo=FALSE, Plot_Three,fig.height = 12, fig.width = 12,fig.align = "center", message=FALSE, warning=FALSE}

ggplot(aes(x = as.factor(quality)), data = redWine.clusters) +
  stat_count(geom = "bar", aes(fill = cluster)) + 
  labs(title = 'Distribution of quality by cluster', x = "quality") +
  theme_tufte()

```

### Description Three

I've chosen this stacked bar chart as my final observation as I think it can 
act as a stepping stone for the next phase of analysis, whether that be refining 
the number of clusters (you get some interesting results using k=5), performing
inference tests or including the variable in a regression model to try to 
predict quality ratings. 

------

## Reflection

This dataset was incredibly tidy and easy to perform standard analysis, a 
perfect playground for learning a new language. I had hoped to learn more about 
red wines in the analysis, but unfortunately without more context it has done 
little to enlighten my next visit to the bottle shop. However, I have learnt a 
lot about the science of wine making, and the chemical reactions to aim for and
avoid when crafting a palatable wine. 

The cluster analysis hints at what are essentially similar chemical profiles.
With some further detail an analyst might be able to infer other properties 
about the wine, such as the region, age or vintage, varieties of grape and 
perhaps even more revealing characteristics about the process of making the 
wine. The reader should note that I haven't carried out a regression analysis 
to predict quality as a matter of principle; personally, I feel that wine is to 
be enjoyed rather than judged.

### A parting quote:
 
> *"Wine can of their wits the wise beguile, Make the sage frolic, and the*
*serious smile"*
                **Homer**

But what does it do to the Data Analyst...?

Thank you for your time.

Guy